Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-02 paper/main.tex
more updates to the paper

In this section we describe our main contribution: Unbalanced Mallows model
(UMM), a novel algorithm for the optimization of an expensive black-box function over
permutation spaces. This algorithm is a probabilistic method based on the
Mallows model and uBorda learning.

\EKHINE{10. In Algorithm 1 'S' and 'F' are *sets* (thus, unordered). Therefore, when adding sigma and f(sigma) the correspondence between the two elements seems to be lost. This mapping would be better represented as a set of *pairs* < sigma, f(sigma) >. You always add these elements to both sets at the same time in step 8, anyway.
}

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, they define the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
\ekhine{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}{}

@@

\subsubsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $\rho \in (0,1)$. %
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$, i.e.,\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
%
\begin{equation}
  \begin{split}
  r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma'\in S'}\rho^{f(\sigma')} \quad\text{where}\; &|S'| = r_2|S| \\[-1.2em]
  &\land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
  \end{split}
\end{equation}
%
i.e., $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

\ekhine{4. Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}{}

\subsubsection{Sampling step}\label{sec:sampling}
In this step, the algorithm samples from the distribution the new data point to be evaluated. 
Sampling from $MM(\sigma_0,\theta)$  can be done efficiently given the parameters  $\sigma_0$ and $\theta$, as studied by \citet{IruCalLoz2016permallows}, where the computational complexity is $O(n\log n)$. % and $O(n^2)$.\MANUEL{X respectively? more than 2}
%
At each iteration of UMM, $\sigma_0=\hat\sigma_0$ is the permutation obtained in the previous learning step. The scale $\theta$ controls the expected value and variance of the distance $D$ of a random permutation $\sigma\sim MM(\sigma_0, \theta)$ to the location parameter $\sigma_0$. Both  the expected value of $D$ and its variance increase monotonically with the value of $\theta$. In particular, the expected distance $\Edist$ is given by~\cite{FliVer1986}:
%
\begin{equation}\label{eq:expectation}
\Edist = \frac{n \cdot \exp(-\theta)}{1-\exp(-\theta)} - \sum_{j=1}^{n}  \frac{j\cdot\exp(-j  \theta)}{1-\exp(-j\theta)}
\end{equation}

@@

\ekhine{5. "At the first iteration, Theta is set so that $E[D] = {n \choose 4}$". Why?}{}
A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}.

@@

\ekhine{The algorithm uses a parameter $\beta$ that determines the number of points sampled, while the authors analyze the sensitivity of the results to the parameters r 1 and r 2, they do not do that with $\beta$. In my opinion the value of $\beta$ is quite relevant.}{}

@@

\subsubsection{CEGO} \ekhine{tenemos una pagina entera dedicada a CEGO :s}{}
Combinatorial Efficient Global Optimization (CEGO)~\citep{ZaeStoFriFisNauBar2014} is an extension of the well-known EGO
method~\citep{JonSchWel98go} to unconstrained black-box combinatorial
optimization problems. In EGO, Gaussian process models are used as a surrogate
of the landscape of the expensive original problem. An optimization method
searches for solutions in the surrogate model by optimizing the expected
improvement criterion, which balances the expected mean and variance of the
chosen solution. Once a solution is chosen, it is evaluated on the actual
fitness function and the result is used to update the surrogate-model,
hopefully increasing its predictive power.

@@

%% MANUEL: Blind, so no acknowledgments yet.
\begin{smaller}
  \paragraph*{Reproducibility.} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
  \vspace*{-1em}
\paragraph*{Acknowledgements.} Blind
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
%\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
%Thanks to: Hao Wang (Leiden University) for pointing us to the arguments of
%\citep{EriPeaGar2019scalable}; Thomas Stützle (Université libre de Bruxelles) for suggesting works
%about differences between QAPLIB instances; and Ahmed Boujaada for the experimental support.

\end{smaller}

>>>>

In this section we describe our main contribution: Unbalanced Mallows model
(UMM), a novel algorithm for the optimization of an expensive black-box function over
permutation spaces. This algorithm is a probabilistic method based on the
Mallows model and uBorda learning.\MANUEL{Perhaps we should call it UMM Algorithm? otherwise people would complain. Or Unbalanced Mallows Algorithm (UMA)?}

\EKHINE{10. In Algorithm 1 'S' and 'F' are *sets* (thus, unordered). Therefore, when adding sigma and f(sigma) the correspondence between the two elements seems to be lost. This mapping would be better represented as a set of *pairs* < sigma, f(sigma) >. You always add these elements to both sets at the same time in step 8, anyway. MANUEL: FIXED
}

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, they define the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. \MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
%
\begin{equation}
  \begin{split}
  r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma'\in S'}\rho^{f(\sigma')} \quad\text{where}\; &|S'| = r_2|S| \\[-1.2em]
  &\land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
  \end{split}
\end{equation}
%
i.e., $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

\subsection{Sampling step}\label{sec:sampling}
In this step, the algorithm samples from the distribution the new data point to be evaluated. 
Sampling from $MM(\sigma_0,\theta)$  can be done efficiently given the parameters  $\sigma_0$ and $\theta$, as studied by \citet{IruCalLoz2016permallows}, where the computational complexity is $O(n\log n)$. % and $O(n^2)$.\MANUEL{X respectively? more than 2}
%
At each iteration of UMM, $\sigma_0=\hat\sigma_0$ is the permutation obtained in the previous learning step. The scale $\theta$ controls the expected value and variance of the distance $D$ of a random permutation $\sigma\sim MM(\sigma_0, \theta)$ to the location parameter $\sigma_0$. Both  the expected value of $D$ and its variance increase monotonically with the value of $\theta$. In particular, the expected distance $\Edist$ is given by~\cite{FliVer1986}:
%
\begin{equation}\label{eq:expectation}
\Edist = \frac{n \cdot \exp(-\theta)}{1-\exp(-\theta)} - \sum_{j=1}^{n}  \frac{j\cdot\exp(-j  \theta)}{1-\exp(-j\theta)}
\end{equation}

@@

\EKHINE{5. "At the first iteration, Theta is set so that $E[D] = {n \choose 4}$". Why?}
A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance   controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}.

@@

\EKHINE{The algorithm uses a parameter $\beta$ that determines the number of points sampled, while the authors analyze the sensitivity of the results to the parameters r 1 and r 2, they do not do that with $\beta$. In my opinion the value of $\beta$ is quite relevant.}

@@

\subsubsection{CEGO} \EKHINE{tenemos una pagina entera dedicada a CEGO :s}\MANUEL{Sugiere qué quitar usando el commando \textbackslash{}ekhine}
Combinatorial Efficient Global Optimization (CEGO)~\citep{ZaeStoFriFisNauBar2014} is an extension of the well-known EGO
method~\citep{JonSchWel98go} to unconstrained black-box combinatorial
optimization problems. In EGO, Gaussian process models are used as a surrogate
of the landscape of the expensive original problem. An optimization method
searches for solutions in the surrogate model by optimizing the expected
improvement criterion, which balances the expected mean and variance of the
chosen solution. Once a solution is chosen, it is evaluated on the actual
fitness function and the result is used to update the surrogate-model,
hopefully increasing its predictive power.

@@

\subsubsection{Reproducibility.} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
%  \vspace*{-1em}
%\begin{acks}
%% MANUEL: Blind, so no acknowledgments yet.
%\paragraph*{Acknowledgements.} Blind
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
%\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
%Thanks to: Hao Wang (Leiden University) for pointing us to the arguments of
%\citep{EriPeaGar2019scalable}; Thomas Stützle (Université libre de Bruxelles) for suggesting works
%about differences between QAPLIB instances; and Ahmed Boujaada for the experimental support. 
  % supported by the \grantsponsor{GS501100001809}{National Natural
  %   Science Foundation of
  %   China}{http://dx.doi.org/10.13039/501100001809} under Grant
  % No.:~\grantnum{GS501100001809}{61273304}
  % and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
  %   Scientists' Support Program}.
%\end{acks}
